---
title: "Analysis of Healthy and Economic Impacts on NOOAA's Storm Database from 1950 - 2011"
author: "Xuecheng LIU"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```
## Synopsis

### The Dataset
This project explores [the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database from 1950 to 2011](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) which tracks characteristics of major storms and weather events in the United States. More detailed information can be found in [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf). 

### The Questions

1. Across the United States, which types of events are most harmful with respect to population health? 
2. Across the United States, which types of events have the greatest economic consequences? 

### The Variables Needed for Analysis

* EVTYPE
* FATALITIES
* INJURIES
* PROPDMG
* PRODMGEXP
* CROPDMG
* CROPDMGEXP

## Data Processing

### Load Data and Packages
This step will download the data if necessary. The `evtyp_list`, as its name indicates, is a vector containing the names of 48 weather events and will be used to justify typos in `EVTYPE`.
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(R.utils)
library(stringdist)

if (!file.exists("repdata%2Fdata%2FStormData.csv")) {
        if (!file.exists("repdata%2Fdata%2FStormData.csv.bz2")) {
                download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                              destfile = "repdata%2Fdata%2FStormData.csv.bz2", method = "curl")
        }
        bunzip2("repdata%2Fdata%2FStormData.csv.bz2", remove = FALSE)
}

storm <- read_csv("repdata%2Fdata%2FStormData.csv")
evtype_list <- read_table("evtype.txt", col_names = FALSE)[[1]]
```

### Subset
Select the 7 needed variables mentioned before.
```{r}
storm <- select(storm, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```
The new `storm` dataset looks like:
```{r}
storm
```

### Clean
1. Make variable names and `evtype_list` to lower case.
```{r}
storm <- mutate(storm, EVTYPE = tolower(EVTYPE))
evtype_list <- tolower(evtype_list)
names(storm) <- tolower(names(storm))
```

2. There are many typos in `evtype` and other events causing more than 48 unique events recorded(nearly a thousand).
```{r}
as_tibble(unique(storm$evtype))
```
`stringdist` package provides `amatch()` function for Approximate String Matching. `evtype_list` is the lookup table for matching. However, some elements that still cannot be matched, such as "summary of June", will return NAs. The strategy for these NAs is to use `drop_na()` to drop rows of `evtype` containing NA.

```{r}
storm$evtype <- evtype_list[amatch(storm$evtype, evtype_list, maxDist = 8)]
storm <- drop_na(storm, evtype)
```
After matching, the unique event numver in `evtype` is perfectly 48.
```{r}
unique(storm$evtype)
```

3. Replace characters in `propdmgexp` and `cropdmgexp` with corresponding exponent values.  

B or b = Billion, M or m = Million, K or k = Thousand, H or h = Hundred. The number from one to ten represent the power of ten (10^The number). The symbols "-", "+" and "?" refers to less than, greater than and low certainty. 

```{r}
storm$propdmgexp <- gsub("[Hh]", "2", storm$propdmgexp)
storm$propdmgexp <- gsub("[Kk]", "3", storm$propdmgexp)
storm$propdmgexp <- gsub("[Mm]", "6", storm$propdmgexp)
storm$propdmgexp <- gsub("[Bb]", "9", storm$propdmgexp)
storm$propdmgexp <- gsub("\\+", "1", storm$propdmgexp)
storm$propdmgexp <- gsub("\\?|\\-|\\ ", "0", storm$propdmgexp)

storm$cropdmgexp <- gsub("[Hh]", "2", storm$cropdmgexp)
storm$cropdmgexp <- gsub("[Kk]", "3", storm$cropdmgexp)
storm$cropdmgexp <- gsub("[Mm]", "6", storm$cropdmgexp)
storm$cropdmgexp <- gsub("[Bb]", "9", storm$cropdmgexp)
storm$cropdmgexp <- gsub("\\+", "1", storm$cropdmgexp)
storm$cropdmgexp <- gsub("\\?|\\-|\\ ", "0", storm$cropdmgexp)

storm$propdmgexp <- as.numeric(storm$propdmgexp)
storm$cropdmgexp <- as.numeric(storm$cropdmgexp)

storm$propdmgexp[is.na(storm$propdmgexp)] <- 0
storm$cropdmgexp[is.na(storm$cropdmgexp)] <- 0
```

## Analysis

### Impact on Population Health
A new variable called `toll`, which equals to `fatalities` plus `injuries` grouped by `evtype`, can be introduced to illustrate impact on population health.
```{r}
storm_to_health <- storm %>%
        select(evtype, fatalities, injuries) %>%
        group_by(evtype) %>%
        summarise(toll = sum(fatalities + injuries)) %>%
        arrange(desc(toll))

knitr::kable(storm_to_health, format = "markdown")
```


### Impact on Economy
Similarly, economic impact can be evaluated by sum of `propdmg` and `cropdmg`. 
```{r}
storm_to_economy <- storm %>% 
        select(evtype, propdmg, propdmgexp, cropdmg, cropdmgexp) %>%
        mutate(propdmg_num = propdmg * (10 ^ propdmgexp), 
               cropdmg_num = cropdmg * (10 ^ cropdmgexp)) %>%
        group_by(evtype) %>%
        summarise(total_damage = sum(propdmg_num + cropdmg_num)) %>%
        arrange(desc(total_damage))
knitr::kable(storm_to_economy, format = "markdown")
```


## Results

According to the two tables above the weather events causing the most population health impact is "tornado", and "flood" the most economic impact.

In addition, the top 10 weather events causing population health and economic impact are: 
```{r}
ggplot(storm_to_health[1:10,], aes(reorder(evtype, toll), toll)) +
        geom_bar(stat = "identity") +
        coord_flip() +
        theme(plot.margin = unit(rep(2,4), "lines")) + 
        labs(title = "Top 10 Weather Events Causing Death and Injury") +
        labs(x = "Event Type", y = "Toll")
```

```{r}
ggplot(storm_to_economy[1:10,], aes(reorder(evtype, total_damage), total_damage)) +
        geom_bar(stat = "identity") +
        coord_flip() +
        scale_y_continuous(labels = scales::dollar) +
        theme(plot.margin = unit(rep(2,4), "lines")) + 
        labs(title = "Top 10 Weather Events Causing Economic Damage") +
        labs(x = "Event Type", y = "Economic Damage")
```







